---
title: "Create a conversion table for the ensembl gene id and hgnc symbol"
author: "Author: Old Vegetable Bird"
date: "`r format(Sys.time(), '%F')`"
output:
    rmarkdown::html_document:
        theme: readable
        highlight: textmate
        df_print: paged
---

# GTF file source

The file [Homo_sapiens.GRCh38.99.gtf](ftp://ftp.ensembl.org/pub/release-99/gtf/homo_sapiens/Homo_sapiens.GRCh38.99.gtf.gz) is the GTF annotation file downloaded from ensembl.org.

# GTF file nine-column specifications
Refers to [GFF/GTF File Format - Definition and supported options](https://uswest.ensembl.org/info/website/upload/gff.html)

1. seqname - name of the chromosome or scaffold; chromosome names can be given with or without the 'chr' prefix. Important note: the seqname must be one used within Ensembl, i.e. a standard chromosome name or an Ensembl identifier such as a scaffold ID, without any additional content such as species or assembly. See the example GFF output below.
2. source - name of the program that generated this feature, or the data source (database or project name)
3. feature - feature type name, e.g. Gene, Variation, Similarity
4. start - Start position of the feature, with sequence numbering starting at 1.
5. end - End position of the feature, with sequence numbering starting at 1.
6. score - A floating point value.
7. strand - defined as + (forward) or - (reverse).
8. frame - One of '0', '1' or '2'. '0' indicates that the first base of the feature is the first base of a codon, '1' that the second base is the first base of a codon, and so on..
9. attribute - A semicolon-separated list of tag-value pairs, providing additional information about each feature.

# GTF file processing

```{r initial-settings}
rm(list = ls())
ptm <- proc.time()
# proc.time() - ptm
options(stringsAsFactors = F)
library("magrittr")
library("readr")
```

# Input files and directories

```{r set-input}
# input:
gtf_download_link <- "ftp://ftp.ensembl.org/pub/release-99/gtf/homo_sapiens/Homo_sapiens.GRCh38.99.gtf.gz"
compressed_gtf_file_name <- "Homo_sapiens.GRCh38.99.gtf.gz"
uncompressed_gtf_file_name <- "Homo_sapiens.GRCh38.99.gtf"
```

# Output files and directories

```{r set-output}
# output:

output_subdirectory <- file.path(getwd(), "intermediate_files") %T>%
  dir.create(., showWarnings = F, recursive = T)

output_base_name <- "ensemble_hgnc_conversion_table_from_gtf"

output_text_file_name <- paste(output_base_name, "txt", sep = ".")
output_compressed_file_name <- paste(output_base_name, "tar.gz", sep = ".")

output_compressed_file_path <- file.path(getwd(), "intermediate_files", output_compressed_file_name)


```

```{r read-input-file}

# GTF file manipulation options
whether_need_download_now <- TRUE
whether_already_compressed <- TRUE
operating_system <- "Windows"

if (whether_need_download_now == TRUE) {
    download.file(url = gtf_download_link,
                  destfile = compressed_gtf_file_name,
                  method = "internal")
}

# You don't have to decompress the gz file to proceed.
# R added transparent decompression for certain kinds of compressed files in the latest version (2.10).
# If you have your files compressed with bzip2, xvz, or gzip,
# they can be read into R as if they are plain text files.
# You should have the proper filename extensions and specify the column types or classes.
if (whether_already_compressed == TRUE) {
    path_to_input_file <- compressed_gtf_file_name
} else {
    path_to_input_file <- uncompressed_gtf_file_name
}

# c stands for character type. Take all 9 columns as character.
column_type_list <- rep("c", times = 9) %>%
    as.list

# First five rows are just descriptions, so we skip them.
raw_gtf_table <- path_to_input_file %>%
    read_delim(file = ., delim = "\t", col_names = F, skip = 5, col_types = column_type_list) %>%
    as.data.frame %>%
    set_colnames(.,  c("seqname", "source", "feature", "start", "end", "score", "strand", "frame", "attribute"))
```


```{r information-extraction}
# Only gene records are interesting.
subset_gtf_table <- raw_gtf_table[raw_gtf_table$feature == "gene", ]

# Extract the ensembl gene ids from the attribute column
ensembl_gene_ids <- subset_gtf_table$attribute %>%
    sapply(., function(x) strsplit(x, ";")[[1]][1]) %>%
    sapply(., function(x) strsplit(x, " ")[[1]][2]) %>%
    gsub("\"", "", ., fixed = TRUE) %>%
    unname

# Extract the hgnc symbols from the attribute column
hgnc_symbols <- subset_gtf_table$attribute %>%
    sapply(., function(x) strsplit(x, ";")[[1]][3]) %>%
    sapply(., function(x) strsplit(x, " ")[[1]][3]) %>%
    gsub("\"", "", ., fixed = TRUE) %>%
    unname

# Match the ensembl gene ids and the hgnc symbols
output_dataframe <- data.frame(ensembl_gene_id = ensembl_gene_ids,
                               hgnc_symbol = hgnc_symbols)
```


```{r write-output-file}
# Output the conversion table of ensembl gene ids and the hgnc symbols.
write.table(output_dataframe, sep = "\t", file = output_text_file_name, quote = F, row.names = F)

# We can compress the output text file into a tar.gz file to save hard disk space.
tar(tarfile = output_compressed_file_name, files = output_text_file_name,
    compression = "gzip", compression_level = 9, tar = "tar")
```


```{r delete_files}
# relocate the compressed file.
file.rename(from = output_compressed_file_name, to = output_compressed_file_path)

# Delete output plain text file. 
unlink(output_text_file_name)

# Delete input gtf file.
unlink(path_to_input_file)

proc.time() - ptm
```

